---
title: "Workshop Contents"
output: html_document
---

This **gt** workshop will teach you *a lot* about what **gt** can do, and, it'll give you an opportunity to experiment with the package. 

The goal of the workshop is to introduce you to a lot of examples and provide some time to generate some sample tables, learning bit-by-bit as we go.

Each module of the workshop focuses on a different family of functions, with relevant examples and exercises in each. They are all presented here as **R Markdown** (.Rmd) files with one file for each workshop module:

- `"01-create-table.Rmd"` (The *Create Table* Family of Functions)

- `"02-create-modify-parts.Rmd"` (The *Create/Modify Parts* Family of Functions)

- `"02z-PRACTICE-01-02.Rmd"` (Practice with examples from modules `01` and `02`)

- `"03-format-data.Rmd"` (The *Format Data* Family of Functions)

- `"04-modify-columns.Rmd"` (The *Modify Columns* Family of Functions)

- `"04z-PRACTICE-03-04.Rmd"` (Practice with examples from modules `03` and `04`)

- `"05-summaries.Rmd"` (The *Add Rows* Family of Functions)

- `"06-table_option_functions.Rmd"` (The Family of *Table Option Functions*)

- `"06z-PRACTICE-05-06.Rmd"` (Practice with examples from modules `05` and `06`)

You can navigate to any of these and modify the code within the self-contained **R Markdown** code chunks. Entire **R Markdown** files can be knit to HTML, where a separate window will show the rendered document.

### The **gt** Installation

Normally you would install **gt** on your system by using `install.packages()`:

```{r eval=FALSE}
# install.packages("gt")
```

For this workshop, however, we are going to use the development version of **gt** and install it from GitHub with `devtools::install_github()`.

```{r eval=FALSE}
# devtools::install_github("rstudio/gt")
```



