---
title: "Creating a Table, Exporting"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gt)
library(tidyverse)
```

## Intro

All **gt** tables begin with a call to `gt()`, where we would supply the input data table (data frame or tibble) and some basic options for creating a stub (`rowname_col`) and row groups (`groupname_col`). We can also introduce a grouped tibble (via **dplyr**’s `group_by()`) for more precise divisions of rows into row groups. 

The `gt_preview()` function is great for getting a nicely-formatted preview of a data table (defaulting to the first 5 rows and the last row).

Finally, sometimes we want to export the table to some format. It could be HTML, it could be RTF... whichever you go with, you can use the `gtsave()` function.

------

### Functions in this module

- `gt()`
- `gt_preview()`
- `gtsave()`

------


### `gt()`: Create a **gt** table object

``` r
gt(
  data,
  rowname_col = "rowname",
  groupname_col = dplyr::group_vars(data),
  process_md = FALSE,
  caption = NULL,
  rownames_to_stub = FALSE,
  row_group_as_column = FALSE,
  auto_align = TRUE,
  id = NULL,
  locale = NULL,
  row_group.sep = getOption("gt.row_group.sep", " - ")
)
```

The `gt()` function creates a **gt** table object when provided with table data. Using this function is the first step in a typical **gt** workflow. Once we have the **gt** table object, we can perform styling transformations before rendering to a display table of various formats.

##### EXAMPLES

The **gt** package contains a few datasets:

- `countrypops` - Yearly populations of countries from 1960 to 2017
- `sza` - Twice hourly solar zenith angles by month & latitude
- `gtcars` - Deluxe automobiles from the 2014-2017 period
- `sp500` - Daily S&P 500 Index data from 1950 to 2015
- `pizzaplace` - A year of pizza sales from a pizza place
- `exibble` - A toy example tibble for testing with **gt**: `exibble`
- `towny` - Populations of all municipalities in Ontario from 1996 to 2021
- `metro` - The stations of the Paris Metro
- `rx_adsl` - An ADSL-flavored clinical trial toy dataset
- `rx_addv` - An ADDV-flavored clinical trial toy dataset

We will use `exibble` frequently during this workshop simply because the entire table easily fits on a screen and contains various types of data (i.e., it's ideal for examples).

Let's print out `exibble` and see what it looks like:

```{r}
exibble
```

Now lets introduce `exibble` to the `gt()` function.

```{r}
exibble |> gt()
```

The above is the basic layout, containing the column labels and the body cells. Numbers are minimally formatted (see the `num` column) but we can apply our own specific formatting (in the next module).

We can modify the layout a bit in the first step and create a stub. A stub is a special column that contains row labels/names. To do this, we supply the name of the column we'd like to use as the stub to the `rowname_col` argument. Conveniently, we have the `"row"` column in `exibble` that's good for this purpose:

```{r}
exibble |> gt(rowname_col = "row")
```

Great! A bit on the *stub*. It doesn't have a column label above it because it's technically not a column. We can still put a label above it though, and this is done with the `tab_stubhead()` function (demo for this in the next module).

We can do some more structuring. Sets of rows can be grouped, showing a *row group label* above each group. We can do this by using a column containing categorical values (usually grouping labels). Conveniently again, `exibble` has the `"group"` column. Let's use that in the `rowname_col` argument of `gt()` and see what it looks like:

```{r}
exibble |>
  gt(rowname_col = "row", groupname_col = "group")
```

Even if rows are in an undesirable order, **gt** will put the rows in the correct group (but still respect the order of rows).

```{r}
exibble |>
  dplyr::sample_n(size = 8) |>
  gt(rowname_col = "row", groupname_col = "group")
```

Say you have a data frame with row names (like `mtcars`). You can put those `row.names` in the stub with `rownames_to_stub = TRUE`:

```{r}
mtcars |>
  dplyr::slice_head(n = 10) |>
  gt(rownames_to_stub = TRUE)
```

If you use **dplyr** then you're probably familiar with the `group_by()` function. This will generate a grouped tibble, and you can pass that into `gt()`. The package will know what to do with a grouped tibble:

```{r}
exibble |> 
  dplyr::group_by(group) |>
  gt(rowname_col = "row")
```

This also works if you have multiple columns grouped. Use `row_group.sep` to specify the separator between labels.

```{r}
pizzaplace |> 
  dplyr::filter(type %in% c("supreme", "veggie")) |>
  dplyr::group_by(type, size) |>
  dplyr::slice_head(n = 3) |>
  gt(row_group.sep = " // ")
```

------

### `gt_preview()`: Preview a **gt** table object

``` r
gt_preview(
  data,
  top_n = 5,
  bottom_n = 1,
  incl_rownums = TRUE
)
```

Sometimes you may want to see just a small portion of your input data. We can use `gt_preview()` in place of `gt()` to get the first *x* rows of data and the last *y* rows of data (which can be set by the `top_n` and `bottom_n` arguments).

##### EXAMPLES

Use `gtcars` to create a **gt** table preview (with only a few of its columns). You'll see the first five rows and the last row.

```{r}
gtcars |>
  dplyr::select(mfr, model, year) |>
  gt_preview()
```

Don't want to see the row numbers? They can be excluded by using `incl_rownums = FALSE`.

```{r}
sp500 |> 
  gt_preview(
    bottom_n = 5,
    incl_rownums = FALSE
  )
```


------

### `gtsave()`: Save a **gt** table as a file

``` r
gtsave(
  data,
  filename,
  path = NULL,
  ...
)
```

The `gtsave()` function makes it easy to save a **gt** table to a file. The function assumes the output file type by the extension provided in the output filename. This will produce either an HTML, PDF, PNG, LaTeX, RTF, or docx file.

##### EXAMPLES

Use `exibble` to create a **gt** table.

```{r}
gt_tbl <-
  exibble |>
  gt(
    rowname_col = "row",
    groupname_col = "group"
  )
```

Write an HTML file.

```{r}
gt_tbl |> gtsave("tab_1.html", inline_css = FALSE)
```

Get a PNG file (essentially a screenshot of the HTML table) by using the .png extension.

```{r}
#gt_tbl |> gtsave("tab_r.png")
```

Get an RTF file by using the .rtf extension in the `filename`.

```{r}
#gt_tbl |> gtsave("tab_r.rtf")
```

------

### SUMMARY

1. To make a **gt** table, get your input table in the right arrangement and use `gt()`.
2. The `gt()` function has a few options for adding group rows and a *stub*.
3. If you need a quick preview of a table (regardless of the table's size) try using `gt_preview()`
4. You can save a **gt** table in a file with `gtsave()`; you have to name the file in a supported format (e.g., `"gt_table.rtf"` for an RTF file containing the table).

