---
title: "Practice with Modules 01 and 02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gt)
library(tidyverse)
```

Practice is good. We learned a great number of functions in modules `01` and `02`. Here they are as a list:

### **gt** functions from module `01`

- `gt()`
- `gt_preview()`
- `gtsave()`

### From module `02`

- `tab_stubhead()`
- `tab_spanner()`
- `tab_header()`
- `tab_source_note()`
- `tab_footnote()`
- `tab_style()`
- `tab_options()`
- `contains()`
- `matches()`
- `starts_with()`
- `ends_with()`
- `everything()`
- `md()`
- `html()`
- `cells_title()`
- `cells_stubhead()`
- `cells_column_spanners()`
- `cells_column_labels()`
- `cells_row_groups()`
- `cells_stub()`
- `cells_body()`
- `cells_summary()`
- `cells_grand_summary()`
- `cell_text()`
- `cell_fill()`
- `cell_borders()`

Although this is a lot, in a short amount of time we can still get useful practice by trying a few things out.

#### Suggestions for practice

Try modifying this statement several times and see what happens (e.g., leave out the first argument, change values, etc.).

```{r}
exibble |> gt(rowname_col = "row", groupname_col = "group")
```

Try adapting this snippet of code to make a great set of `title` and `subtitle`. Use `md()` to get Markdown in there. Know HTML? Try your hand at some HTML titling with the `html()` helper. Use `gtsave()` to export the table to your preferred output type.

```{r}
gtcars |>
  dplyr::select(mfr, model, msrp) |>
  dplyr::slice(1:5) |>
  gt() |>
  tab_header(
    title = "<title>",
    subtitle = "<subtitle>"
  )
```

To add styles in a **gt** table, we need to use the combination of `tab_style()`, one or more location helper functions (e.g,. `cells_body()`, `cells_column_labels()`, etc.), and a styling function like `cell_fill()`. Here's an example table with a number of different location types. Try changing the background of different locations.

```{r}
exibble |>
  gt(rowname_col = "row", groupname_col = "group") |>
  tab_header(title = "The title", subtitle = "The subtitle") |>
  tab_source_note("A source note.") |>
  tab_footnote(
    footnote = "A footnote",
    locations = cells_body(columns = 1, rows = 1)
  ) |>
  tab_style(
    style = cell_fill(color = "#E5873A"),
    locations = cells_body(columns = char, rows = "row_1")
  )
```

