---
title: "Practice with Modules 05 and 06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gt)
library(tidyverse)
```

Final practice session. What a long and winding road it's been.

### **gt** functions from module `05`

- `summary_rows()`
- `grand_summary_rows()`

### From module `06`

- `opt_footnote_marks()`
- `opt_row_striping()`
- `opt_align_table_header()`
- `opt_all_caps()`
- `opt_table_lines()`
- `opt_table_outline()`
- `opt_table_font()`
- `opt_css()`
- `google_font()`
- `default_fonts()`
- `info_google_fonts()`

#### Suggestions for practice

Try out some of the `opt_*()` functions on the table below. Also, set a default table font using `opt_table_font()` and `google_font()`. Use `info_google_fonts()` to get the names of some of Google's Fonts.

```{r}
exibble %>%
  gt(rowname_col = "row", groupname_col = "group") %>%
  tab_header(title = "The title", subtitle = "The subtitle") %>%
  tab_source_note("A source note.") %>%
  tab_footnote(
    footnote = "A footnote",
    locations = cells_body(columns = 1, rows = 1)
  ) %>%
  tab_style(
    style = cell_fill(color = "#E5873A"),
    locations = cells_body(columns = char, rows = 1)
  )
```

